---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Henry Yang"
date: 09/24/2025
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Alaska Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)

# Set your Census API key
census_api_key(Sys.getenv("a429b978d14aa85d33761ac9dead46b91ad63915"), install = TRUE, overwrite = TRUE)

# Choose your state for analysis - assign it to a variable called my_state
state_choice <- "AK"
```

**State Selection:** I have chosen **Alaska** for this analysis because: Far away from the continental america, beautiful and mysterious, tempting me to explore it.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
ak_demo <- get_acs(
  geography = "county",
  variables = c(
    med_inc = "B19013_001",
    tot_pop = "B01003_001"
  ),
  state = "AK",
  year = 2022,
  output = "wide"
) %>%
  mutate(county_name = str_remove(NAME, ", Alaska"))

# Display the first few rows
head(ak_demo)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
ct_reliability <- ak_demo %>%
  mutate(
    medinc_moe_pct = (med_incM / med_incE) * 100,
    reliability_cat = case_when(
      medinc_moe_pct < 5 ~ "High Confidence (<5%)",
      medinc_moe_pct <= 10 ~ "Moderate Confidence (5~10%)",
      TRUE ~ "Low Confidence (>10%)"
    ),
    unreliable_flag = case_when(
      medinc_moe_pct > 10 ~ "🔴"
    )
  )

# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
reliability_summary <- ct_reliability %>%
  count(reliability_cat) %>%
  mutate(pct = round(100 * n / sum(n), 1))

reliability_summary
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
high_uncertainty <- ct_reliability %>%
  arrange(desc(medinc_moe_pct)) %>%
  slice(1:5) %>%
  select(NAME, med_incE, med_incM, medinc_moe_pct, reliability_cat)

# Format as table with kable() - include appropriate column names and caption
kable(
  high_uncertainty,
  col.names = c("County", "Median Income", "Margin of Error", "MOE %", "Reliability"),
  caption = "Top 5 Alaska Counties with Highest Median Income MOE Percentage",
  format.args = list(big.mark = ","),
  digits = 1
)
```


**Data Quality Commentary:** These results suggest that counties such as Dillingham, Yakutat, and Haines could be poorly served by algorithms relying on median income data, since their estimates have very high uncertainty. In small or remote populations like these Alaska counties, limited sample sizes could inflate margins of error. As a result, algorithms that assume the data is precise may reinforce misallocated resources in these areas.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_ct <- ct_reliability %>%
  filter(NAME %in% c("Anchorage Municipality, Alaska",
                     "Fairbanks North Star Borough, Alaska",
                     "North Slope Borough, Alaska")) %>%
  select(NAME, tot_popE, med_incE, medinc_moe_pct, reliability_cat)

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
kable(
  selected_ct,
  col.names = c("County", "Total Population", "Median Income", "MOE %", "Reliability"),
  caption = "Selected Alaska Counties with Highest Median Income MOE Percentage",
  format.args = list(big.mark = ","),
  digits = 1
)
```


**Comment on the output:** The output shows how population size often aligns with data reliability: Anchorage, with nearly 300,000 people, has a very low MOE (3%) and thus highly reliable income data. Fairbanks, with a mid-sized population, has moderate reliability, while the sparsely populated North Slope shows much higher uncertainty (21.4%), making its income estimates less dependable for policy or algorithmic use.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide

- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
races <- c(
  white = "B03002_003",
  black = "B03002_004",
  hispanic = "B03002_012",
  total = "B03002_001"
)

tract_code <- c("020", "090", "185")

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
tract_demo <- get_acs(
  geography = "tract",
  state = "AK",
  county = tract_code,
  year = 2022,
  variables = races,
  output = "wide"
)

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
# Add readable tract and county name columns using str_extract() or similar
tract_demo_pct <- tract_demo %>%
  mutate(
    pct_white = 100 * whiteE / totalE,
    pct_black = 100 * blackE / totalE,
    pct_hispanic = 100 * hispanicE / totalE,
    tract_name = str_extract(NAME, "Census Tract [0-9.]+"),
    county_name = NAME %>%
                  str_remove("Census Tract [0-9.]+; ") %>%
                  str_remove("; Alaska")
  )

tract_demo
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
high_hispanic <- tract_demo_pct %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(tract_name, county_name, pct_hispanic, hispanicE, totalE)

high_hispanic

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
ct_avg_demo <- tract_demo_pct %>%
  group_by(county_name) %>%
  summarize(
    n_tracts = n(),
    avg_pct_white = mean(pct_white, na.rm = TRUE),
    avg_pct_black = mean(pct_black, na.rm = TRUE),
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE)
  ) %>%
  arrange(desc(n_tracts))

# Create a nicely formatted table of your results using kable()
kable(
  ct_avg_demo,
  col.names = c("County", "# Tracts", "White %", "Black %", "Hispanic %"),
  caption = "Average Racial/Ethnic Composition by County (Tract-Level Averages)",
  digits = 1
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
tract_reliability <- tract_demo %>%
  mutate(
    white_moe_pct = (whiteM / whiteE) * 100,
    black_moe_pct = (blackM / blackE) * 100,
    hispanic_moe_pct = (hispanicM / hispanicE) * 100,
    unreliable_flag = case_when(
      white_moe_pct > 15 | black_moe_pct > 15 | hispanic_moe_pct > 15 ~ "🔴"),
    tract_name = str_extract(NAME, "Census Tract [0-9.]+"),
    county_name = NAME %>%
                  str_remove("Census Tract [0-9.]+; ") %>%
                  str_remove("; Alaska")
  )

# Create summary statistics showing how many tracts have data quality issues
ct_moe_summary <- tract_reliability %>%
  group_by(county_name) %>%
  summarize(
    total_tracts = n(),
    unreliable_tracts = sum(!is.na(unreliable_flag))
  )

ct_moe_summary
```

**Conclusion:** All of the tracts are unreliable!

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
## Since all of the tracts have high MOE issues, I will be only using high MOE of white only population for analysis.
# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison
race_summary <- tract_reliability %>%
  mutate(high_white_moe = white_moe_pct > 15) %>%
  group_by(high_white_moe) %>%
  summarize(
    n_tracts = n(),
    avg_population = mean(totalE, na.rm = TRUE),
    avg_pct_white = mean((whiteE / totalE)[is.finite(whiteE / totalE)], na.rm = TRUE) * 100,
    avg_pct_black = mean((blackE / totalE)[is.finite(blackE / totalE)], na.rm = TRUE) * 100,
    avg_pct_hisp = mean((hispanicE / totalE)[is.finite(hispanicE / totalE)], na.rm = TRUE) * 100
  )

# Create a professional table showing the patterns
kable(
  race_summary,
  col.names = c("Unreliable Tracts", "# Tracts", "Average Population", "White %", "Black %", "Hispanic %"),
  caption = "Comparison of Tracts by White Alone MOE Reliability",
  digits = 1
)
```


**Pattern Analysis:** Tracts with unreliable White MOE are much more common (74 vs. 12), and they tend to have slightly smaller populations on average. These unreliable tracts also show less White residents and somewhat more Hispanic population. This suggests that data quality problems are concentrated in smaller or more diverse tracts, where ACS sample sizes are limited and subgroup estimates become less precise.

# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**

1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?

2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?

3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?

4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?


**Executive Summary:**

**Overall Pattern Identification**
Across Alaska counties, more than half (53.3%) of median household income estimates fall into the “Low Confidence” category, while only 10% qualify as “High Confidence.” The top 3 counties with the highest MOE percentages, such as Dillingham, Yakutat, and Haines, show margins of error above 20%. This indicates systematic reliability challenges in smaller and remote communities. At the tract level, Anchorage demonstrates greater diversity and larger populations, while North Slope shows both lower White representation (32.6%) and smaller sample sizes, reinforcing the vulnerability of rural areas to unreliable data.

**Equity Assessment**
Communities most at risk of algorithmic bias are those in rural and remote counties. For example, Dillingham and Yakutat would be poorly served by algorithms relying on median income, as their estimates carry uncertainty levels of 30–40%. At the tract level, unreliable tracts (74 out of 86) tend to have slightly smaller populations and lower White percentages compared to reliable tracts, suggesting that diverse or minority-heavy communities face greater risks of data-driven misrepresentation.

**Root Cause Analysis**
ACS sample sizes are smaller in rural and low-population tracts, which inflates MOE and decreases estimate stability. This might because Geographic isolation compounds these problems by making data collection more difficult. This indicates that data quality problems are not randomly distributed, but systematically concentrated in communities that are already more vulnerable.

**Strategic Recommendations**
The Department should prioritize three measures to strengthen equity and reliability in data-driven policy. First, apply MOE thresholds in all algorithmic workflows to flag unstable estimates before use. Second, conduct equity audits to ensure algorithms do not misallocate resources to rural or minority communities. Third, communicate uncertainty transparently so policymakers understand the limitations of ACS data in Alaska’s rural and remote contexts.

## 5.2 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category
ct_reliability_sum <- ct_reliability %>%
  select(NAME, med_incE, medinc_moe_pct, reliability_cat)

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"
ct_recommendation <- ct_reliability_sum %>%
  mutate(
    Recommendation = case_when(
      reliability_cat == "High Confidence (<5%)" ~ "Safe for algorithmic decisions",
      reliability_cat == "Moderate Confidence (5~10%)" ~ "Use with caution - monitor outcomes",
      reliability_cat == "Low Confidence (>10%)" ~ "Requires manual review or additional data"
    )
  )

# Format as a professional table with kable()
kable(
  ct_recommendation,
  col.names = c("County", "Median Income", "MOE %", "Reliability", "Recommendation"),
  caption = "Algorithm Implementation Framework by County for Alaska",
  format.args = list(big.mark = ","),
  digits = 1
)
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:**
 - Anchorage Municipality
 - Kenai Peninsula Borough
 - Matanuska-Susitna Borough
These 3 counties have highly reliable estimates (margins of error below 5%), making it suitable for immediate algorithmic decision-making.

2. **Counties requiring additional oversight:**
 - Aleutians East Borough
 - Aleutians West Census Area
 - Fairbanks North Star Borough
 - Hoonah-Angoon Census Area
 - Juneau City and Borough
 - Ketchikan Gateway Borough
 - Kusilvak Census Area
 - Nome Census Area
 - Northwest Arctic Borough
 - Prince of Wales-Hyder Census Area
 - Sitka City and Borough
These 11 counties may have algorithms being used here, but outputs should be closely monitored through periodic audits, error tracking, and targeted human review in high-stakes contexts.

3. **Counties needing alternative approaches:**
 - Bethel Census Area
 - Bristol Bay Borough
 - Chugach Census Area
 - Copper River Census Area
 - Denali Borough
 - Dillingham Census Area
 - Haines Borough
 - Kodiak Island Borough
 - Lake and Peninsula Borough
 - North Slope Borough
 - Petersburg Borough
 - Skagway Municipality
 - Southeast Fairbanks Census Area
 - Wrangell City and Borough
 - Yakutat City and Borough
 - Yukon-Koyukuk Census Area
These 16 counties (over half of Alaska’s total) show margins of error exceeding 10%, with some reaching above 30%. Automated decision-making here risks significant misallocation. The Department should consider manual review, multi-year or regional aggregation, or integration of supplemental state/tribal data before using these estimates.

## 5.3 Questions for Further Investigation
1.	Are unreliable estimates spatially concentrated in rural or Arctic regions, and how does geographic isolation contribute to these patterns?
2.	Do particular demographic groups like Alaska Native consistently face higher MOE values, and what adjustments are needed to ensure equity in algorithmic use?

# Technical Notes

**Data Sources:** 

 - U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
 
 - Retrieved via tidycensus R package on 09-24-2025

**Reproducibility:** 

 - All analysis conducted in R version 2025.05.1+513
 
 - Census API key required for replication
 
 - Complete code and documentation available at: https://musa-5080-fall-2025.github.io/portfolio-setup-henryyzh1/

**Methodology Notes:**

 - Analysis is based on the 2022 ACS 5-year estimates for B19013_001, B01003_001 in county level, B03002_003, B03002_004, B03002_012, B03002_001 in census tract level
 
 - Analysis is based on R packages like tidycensus, tidyverse, knitr
 
 - Reliability categories were defined by MOE percentage: High (<5%), Moderate (5–10%), and Low (>10%)
 
 - County selection is based on reliability category and total population

**Limitations:**
 - Sample size constraints: There are many rural counties in Alaska have small populations, leading to inflated MOEs and unstable estimates
 
 - Temporal scope: Results are based on a single year (2022); year-to-year variability is not captured
 
 - Applicability: The MOE thresholds provide a general framework, but tolerance for uncertainty may differ depending on policy area

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`